Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-03-31 doc/paper.Rmd
Working on discussion.

The mean proportion of flowering plants across all census years was generally higher in continuous forest (0.050 ± 0.021) compared to 1-ha fragments (0.036 ± 0.027).
In fact, the proportion of flowering plants was higher in continuous forest in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
The most striking difference between habitat types in flowering proportion coincided with a severe drought in 2003 when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest and 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
<!--# not sure how necessary this is -->

@@

@@

-   @dowd2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production by Heliconia (one step past flaowering, so might be worth citing to say that there are other ways reprod. can be affected

@@

-   Many humid tropical forest species time reproduction to coincide with dry season peak to avoid low irradiance in wet season.
    (refs within @wrightSeasonalNinoLonger2006. van Shaik et al. 1993; Wright and vanSchaik 1994).
    So rather than bet-hedging, maybe extra dry dry seasons are actually just really prime conditions for *Heliconia* reproduction and recruitment.

-   \

@@

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].

This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
We found evidence for delayed effects of drought which, in our models, are sometimes stronger predictors of vital rates than more recent conditions.
Distributed lag non-linear models offer an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015a for alternative methods), yet few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].

In addition to finding delayed effects of SPEI on demography, we found that these effects differed between forest fragments and continuous forest habitat.
Overall survival of plants was similar in fragments and continuous forest for all plant sizes.
Although plants were smaller in forest fragments, they grew slightly faster than plants in continuous forest, closing the size gap over the decade of demographic data.
However, at all sizes, variation in SPEI had a greater impact on both growth and survival in fragments compared to continuous forest.
Both unusually wet and unusually dry conditions resulted in reductions in survival in fragments, with lagged effects of up to around 15 months.
Similarly for plant size, extremes in both directions in the census month caused reductions in growth in fragments.
However, the lagged effects of SPEI on plant size depended on season with the general pattern of extremes (unusually wet wet seasons and unusually dry dry seasons) being detrimental to plant growth.

Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.
Average flowering probability of plants in continuous forest is equivalent to flowering once every 125 years compared to once every 288 years in 1-ha fragments (assuming average plant size and SPEI).
<!--# this doesn't make sense.  Would be better to pick a large plant size and plug in to both --> We also see delayed effects of drought in flowering with increased flowering in years when plants experienced a drought two dry seasons ago (15--20 months prior to census).

-   What look like small changes in vital rates add up over time, especially for slow-growing, long-lived organisms [@gagnonGrowthUnderstoryHerb2011].

-   Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability.

1.  Mechanisms [this lays out the biological, fragmentation-related mechanisms underpinning these patterns]

    -   I gather that in a good year, plants can store carbon and nutrients in rhizomes and then mobilize as needed and the opposite is true in a bad year.
        Flowering is a big investment so perhaps flowering in one year means that you don't flower the following one.
        If that is the case, it would make sense to include growth in the previous year as a covariate in all analyses rather than do this indirectly via a potential effect of past SPEI on past growth.
        It will also be easier to explain as you write the paper.
        **Emilio will look into physiological responses to light and H20 in the horticultural and ecological literature, eg. Westerband papers, esp if focusing on herbs** Basically anything on how more or less precip influences vital rates or how plants respond to them, esp if linked to changes in canopy openness, litterfall, flowering in ENSO years, Bruna transplant experiments, Heliconia germination in secondary forest, important to emphasize: results are often counter-intuitive.

    -   cite Esteban et al. 2021 for negative effects of large positive SPEI

    -   The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation. e.g. S. Joe Wright trees flowering after ENSO -->.

    -   Alternatively, the delayed effect of drought could be due to indirect effects mediated by canopy openness.
        In Manaus, canopy is as green or greener in dry season compared to wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought has the potential to increase solar radiation if it causes reductions in canopy cover.
        Tree mortality (CITATIONS) and litterfall (CITATIONS) in the Amazon increase following drought, which could increase solar radiation to the understory resulting in greater energy capture by *H. acuminata*.

    -   We also considered that some delayed effects could be a result of a cost of reproduction (triggered by an extreme SPEI) leading to reduced growth, survival, and/or flowering probability in the following season.
        However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
        <!--# double-check against models and edit. -->

2.  Ecological Consequences of these patterns: an eye towards a drier/wetter future

-   More work to be done on flowering responses to drought x light x season interactions.

-   Address how all but the smallest sizes are shrinking on average over the period of this study (Figure \@ref(fig:growth-size)).
    Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.

6.  Buffering [where does this fit best...as a mechanism or something we need to pin down with more work?]

@@

Caveats:

<!--# need to edit.  Moved from methods. -->

One limitation of the DLNM approach is that it does not currently allow for modeling of interactions with a crossbasis function.
This limitation meant that we were unable to model a potential size by environment (i.e. drought history) interaction, which may occur with many plant species [@tredennickSizebyenvironmentInteractionsNeglected2018].
To check for a size by environment interaction, we followed the methods of Tredennick et al. [-@tredennickSizebyenvironmentInteractionsNeglected2018] and fit a pair of generalized linear mixed effects models (GLMMs) for each vital rate using the *lme4* package [@batesFittingLinearMixedeffects2015].
Rather than explicitly modeling the effects of drought history, we modeled environmental stochasticity with random effects of year.
For each vital rate, we essentially replaced the crossbasis function in Eq 1.
with a random effect of year on either the intercept or both the intercept and relationship with plant size.
We assessed the significance of the year-by-size interaction by testing the likelihood ratio of models with and without the interaction against an expected likelihood ratio distribution under the null hypothesis of no interaction [@tredennickSizebyenvironmentInteractionsNeglected2018].

@@

-   There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartzFragmentationForestStructure2019]).
    Smaller/larger plants might be more susceptible to drought.
    So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.

>>>>

There was a pronounced threshold in flowering probability corresponding with a log size of about 5 (Figure \@ref(fig:flwr-size)).
The mean proportion of flowering plants across all census years was generally higher in continuous forest (0.050 ± 0.021) compared to 1-ha fragments (0.036 ± 0.027).
In fact, the proportion of flowering plants was higher in continuous forest in all survey years except 2006 and 2008 (Figure \@ref(fig:eda)c).
The most striking difference between habitat types in flowering proportion coincided with a severe drought in 2003 when the percentage of flowering plants with four or more shoots was 24.1% in continuous forest and 12% in 1-ha fragments (Figure \@ref(fig:eda)c).
The back-transformed model intercepts, representing the flowering probability of averaged sized plants in each habitat given average SPEI, showed a similar pattern with 0.008 for continuous forest and 0.004 for 1-ha fragments.
<!--# not sure how necessary this is -->

@@

<!--# instead of rug plot, add second panel or margin with box plots or violin plots of size at both habitats -->

@@

-   @dowd2020 (unpublished thesis, could write MB and ask for update): precipitation and fragmentations influenced fruit/seed production (conditional on flowering) by *Heliconia* (one step past flowering, so might be worth citing to say that there are other ways reprod. can be affected

@@

-   Many humid tropical forest species time reproduction to coincide with dry season peak to avoid low irradiance in wet season.
    (refs within @wrightSeasonalNinoLonger2006. van Shaik et al. 1993; Wright and vanSchaik 1994).
    But *Heliconia* times reproduction so seeds only germinate at start of rainy season.

-   cite Esteban et al. 2021 for negative effects of large positive SPEI\

@@

Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].
<!--# more here -->

This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
We found evidence for delayed effects of drought which are sometimes stronger predictors of vital rates than more recent conditions.
The approach we took offers an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015a for alternative methods), yet few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].
This demonstrates the importance of long term demographic experiments in determining the effects of changing habitats on populations.

```{r flwr-prob-quartile, include=FALSE}

#figure out flowering probability of largest quartile of plants in each habitat
tar_load(f_cf)
tar_load(f_1ha)

newdf_cf <-
  f_cf$model %>% #use model data frame
  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
  filter(quartiles == 4)

newdf_1ha <-
  f_1ha$model %>% 
  mutate(quartiles = ntile(log_size_prev, 4)) %>% 
  filter(quartiles == 4)

#equivalent to flowering once every __ years
1/predict(f_cf, newdata = newdf_cf) %>% mean() %>% plogis()
1/predict(f_1ha, newdata = newdf_1ha) %>% mean() %>% plogis()
```

*Heliconia acuminata* individuals were smaller in forest fragments, although they grew slightly faster than plants in continuous forest, closing the size gap over the decade of demographic data.
Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002], showing that probability of flowering increases with plant size and plant growth decreases with plant size <!--# other citations? -->. Flowering probability for the largest plants is actually higher in forest fragments; however, because there are fewer large plants in fragments overall flowering probability is much lower. The average fitted probability of flowering for plants in the upper quartile for size is 0.125 in continuous forest and 0.094 in forest fragments, equivalent to flowering once every 7.97 years and 10.63 years, respectively. Seemingly small differences in vital rates such as these can be extremely important as they compound over many years, especially in long-lived organisms[@gagnonGrowthUnderstoryHerb2011].

We found that the effects of variation in SPEI differed between forest fragments and continuous forest habitat.
At mean SPEI, overall survival of plants was similar in fragments and continuous forest, but extremes in SPEI, both positive (wet) and negative (dry) had much greater detrimental effects on survival in habitat fragments.
Similarly variation in SPEI had a greater impact on plant size in fragments compared to continuous forest with a general trend of unusually wet wet seasons and unusually dry dry seasons having a greater negative impact in fragments.
Flowering probability showed a different pattern with a greater effect of variation in SPEI in continuous forests.
However, with few plants large enough to flower, it is likely we simply did not detect the full effects of variation in SPEI on flowering probability in forest fragments with our models.
Both variation in precipitation and extreme conditions are predicted to increase as climate change progresses ( <!--# cite IPCC? -->) and our results demonstrate that this may have a greater negative impact on understory plants in forest fragments compared to continuous forests.

1.  Buffering [where does this fit best...as a mechanism or something we need to pin down with more work?]

@@

Past climate conditions were important for survival, size, and flowering probability.
Extremes in SPEI in the 15 months prior to the census reduced survival in forest fragments (Figure \@ref(fig:surv-spei)).
Extremes in SPEI in the past 36 months affected plant size, although the direction of the effects was seasonally dependent (Figure \@ref(fig:growth-spei)).
Generally, unusually dry dry seasons and unusually wet wet seasons had negative impacts on plant size, and this pattern was more extreme in fragments compared to continuous forest.
For flowering, the SPEI of the current wet season and the dry season two dry seasons ago (lag of 15--20 months) had the strongest influence on flowering probability with dry weather corresponding to an increase in flowering (Figure \@ref(fig:flwr-spei))<!--# wondering if this is more of a results paragraph? Alternatively it goes *after* the next paragraph and includes more speculation about why each specific delayed effect exists.  Might end up integrating with a later paragraph. -->.

There are several possible non-mutually-exclusive explanations for delayed effects of SPEI on vital rates.
First, delayed effects may represent a demographic delay since we do not know precisely when since the previous survey the plant died or changed size.
For example, if extreme drought in the dry season before the census increases plant mortality in that season, this would show up as a delayed effect in our models (e.g. in Figure \@ref(fig:surv-spei)b) despite the impact of drought potentially being immediate.
This explanation for delayed effects applies only to plant size and survival, as flowering occurs very close to the yearly demographic census (Jan--Apr).

Second, these delayed effects may represent a physiological delayed effect of environmental conditions. Under horticultural conditions, flowering shoots of *Heliconia chartacea* that flower in Spring begin to emerge in June or July of the previous year, and the developing inflorescence may be aborted in response to adverse conditions up to 6 months after initiation [@criley1994]. <!--# more here about general trends in physiological delays for plants witht tubers or rhizomes -->Although flowering occurs in late January through April in *H. acuminata*, the initiation of a primordial flowering shoot <!--# not sure this is the right term.  "Differentiation of the apical meristem" ? --> likely happens many months earlier. The increase in flowering rate of plants following a drought could also be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018] <!--# add another more general citation. e.g. S. Joe Wright trees flowering after ENSO -->.
We also considered that some delayed effects could be a result of a cost of reproduction leading to reduced growth, survival, and/or flowering probability in the following season.
However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
Our models show that drought two dry seasons prior to the census increase flowering probability.
A greater understanding of the effects of light, drought, season, and their interactions on flowering in *H. acuminata* is needed to provide a probable mechanism for this pattern.

Finally, delayed effects could be mediated by indirect effects and ecological interactions.
<!--# I'm not sure if this is really an explanation for the *delay* per se.  Need to think about this. -->Drought in tropical forests can increase tree mortality and leaf drop, resulting in decreased canopy cover and increased light penetration (<!--# citations -->).
In Manaus, the canopy is as green or greener in the dry season compared to the wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought may result in relatively large increases in light penetration to the understory.
This increase in light could then result in increased growth or flowering for understory plants ( <!--# citations -->).
<!--# Heliconia grow better at edges, where there's more light -->Here, the delayed effect could be a combined delay of the response of canopy trees to drought and the response of *H. acuminata* to increased insolation.

-   Ecological Consequences of these patterns: an eye towards a drier/wetter future

-   Address how all but the smallest sizes are shrinking on average over the period of this study <!--# this is not true.  A bug previously made it look like all plants were shrinking.  Keeping this because citations -->(Figure \@ref(fig:growth-size)).
    Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.

@@

-   There could be size-specific effects: Fragmentation affects plant size (in *Heliconia* [@brunaDemographicEffectsHabitat2005] and trees [@schwartzFragmentationForestStructure2019]).
    Smaller/larger plants might be more susceptible to drought.
    So differential effects of drought in fragments and continuous forest could be due to an interaction between drought and plant size.
    The `dlnm` package does not currently allow for modeling of interactions, and we lacked data on plant size at the same time scale as SPEI (monthly), so interactions with size were not included in this analysis.
